Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2024-01-05 analysis/paper/paper.Rmd
Added link to new version of Zenodo archive
Updates before PCI recommendation

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.8373857. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

# Declaration of competing interest{-}
The authors have no competing interests to declare.

@@

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  # \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.10449324. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

# Acknowledgement{-}
We would like to thank David Wright for commenting on an earlier draft of this paper. We would also like to thank the reviewers for their constructive feedback and PCI archaeology for facilitating their evaluation of our manuscript.

# Funding{-}
This research received no external funding.

# Conflict of interest disclosure{-}
The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article.

# Data, scripts, code, and supplementary information availability{-}
All data, code, text and figures used for this paper is freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c. This has been archived on Zenodo at https://doi.org/10.5281/zenodo.10449324.

@@

# Supplementary material{-}
